---
title: "558 Homework 3"
author: "Susan Hajmohammad"
format: pdf
editor: visual
---

```{r}

library(tidyverse)
library(palmerpenguins)

```

##Task 1

### Question a:

```{r}

data_a <- read_csv("data.txt")

```

We can't use read_csv() function for reading in this dataset, because the delimiator for that function is flat-file, comma separated or tab separated.The function we can use is read_csv2() because it uses the ";" as the separator which is what we have in our dataset.

```{r}
data <- read_csv2("data.txt")
data
```

### Question b

```{r}
data2 <- read_delim("data2.txt", delim = "6", col_types = "fdc")
data2
```


## Task 2

### Question a

```{r}

trailblazer <- read.csv("trailblazer.csv")
glimpse(trailblazer)

```
### Question b

```{r}

trailblazer_long <- trailblazer |> 
  pivot_longer(cols = Game1_Home:Game10_Home, 
               names_to = c("Game", "Location"),
               names_sep = "_",
               values_to = "Points") 

trailblazer_long
 
```

### Question c

```{r}

trailblazer_wide <- trailblazer_long |> 
  pivot_wider(names_from = Location,
        values_from = Points)|>
  group_by(Player)|>
  mutate(mean_home = mean(Home, na.rm = TRUE))|>
  mutate(mean_away = mean(Away, na.rm = TRUE))|>
  mutate(diff = mean_home - mean_away)|>
  arrange(desc(diff))


trailblazer_wide

```
On average Jusuf Nurkic score more points at home than away in the first 10 games. 

## Task 3
### Question a 

```{r}
library(palmerpenguins)
data(penguins)
head(penguins)

penguins |>
select(species, island, bill_length_mm) |>
pivot_wider(
names_from = island, values_from = bill_length_mm
)

```
What is going on here: Using pivot_wider() here is complicated because there are duplicate penguins that are the same species per island so it doesn't which bill length to use for each species. 

<list> means that for each column, instead of there being a single value for each penguin species, there is instead a list of values in each cell. 

<dbl 52> here means that for the Adelie species instead of being one number it's actually a vector of 52 values.  

<null> means there weren't any of that penguin species on that island, so there are no values for that cell. 

### Question b

```{r}

penguins |>
select(species, island, bill_length_mm) |>
  count(species, island) |>
pivot_wider(names_from = island, 
      values_from = n) 

```

## Task 4

### Question a

Replace bill length Adelie NA with 26.  Replace bill length Gentoo NA with 30. 

```{r}


  

penguins |> 
  #select(species, island, bill_length_mm)|>
  mutate(bill_length_mm = 
           case_when(is.na(bill_length_mm) & species == "Adelie" ~ 26,
         is.na(bill_length_mm) & species == "Gentoo" ~ 30,
         .default = bill_length_mm))|>
  arrange(bill_length_mm) |> 
  slice(1:10)
  


```









