---
title: "558 Homework 3"
author: "Susan Hajmohammad"
format: pdf
editor: visual
---

```{r}

library(tidyverse)
library(palmerpenguins)

```

##Task 1

### Question a:

```{r}

data_a <- read_csv("data.txt")

```

We can't use read_csv() function for reading in this dataset, because the delimiator for that function is flat-file, comma separated or tab separated.The function we can use is read_csv2() because it uses the ";" as the separator which is what we have in our dataset.

```{r}
data <- read_csv2("data.txt")
data
```

### Question b

```{r}
data2 <- read_delim("data2.txt", delim = "6", col_types = "fdc")
data2
```


## Task 2

### Question a

```{r}

trailblazer <- read.csv("trailblazer.csv")
glimpse(trailblazer)

```
### Question b

```{r}

trailblazer_long <- trailblazer |> 
  pivot_longer(cols = Game1_Home:Game10_Home, 
               names_to = c("Game", "Location"),
               names_sep = "_",
               values_to = "Points") 

trailblazer_long
 
```

### Question c

```{r}

trailblazer_wide <- trailblazer_long |> 
  pivot_wider(names_from = Location,
        values_from = Points)|>
  group_by(Player)|>
  mutate(mean_home = mean(Home, na.rm = TRUE))|>
  mutate(mean_away = mean(Away, na.rm = TRUE))|>
  mutate(diff = mean_home - mean_away)|>
  arrange(desc(diff))


trailblazer_wide

```
On average Jusuf Nurkic score more points at home than away in the first 10 games. 











